Characterizing Stellar Streams by ML

### 2023-08-29 src/tex/ms.tex
[showyourwork] overleaf sync

    We generate synthetic stream observations by sampling uniformly in $\phi_1$
    In $\phi_2$ we sample from a quadratic in $\phi_1$ with normally-distributed
    $0.15^{\circ}$ scatter.  The distances are sampled from a linear funciton in
    $\phi_1$ with normally-distributed $0.25 \unit{kpc}$ scatter, then
    transformed to the parallax $\parallax$.  Non-poissonian density variations
    are then simulated by evaluating the stream density with two Gaussian
    distributions centered at different values of $\phi_1$ with random
    amplitudes (which sum to $1$) and widths.  Stream stars are then removed
    using random sampling, weighted by their likelihood of belonging to the
    density. The background astrometric points in are generated from exponential
    distributions at each $\phi_1$.  The number of stream stars is
    $\variable{output/mock/nstream_variable.txt}$, while the number of
    background stars is $\variable{output/mock/nbackground_variable.txt}$.

    The synthetic stream's photometry is also simulated using a 12~Gyr, [Fe/H] =
    -1.35~dex NMIST isochrone \TODO{[cite brutus]}. The isochrone truncates just
    short of the giant branch, and is sampled from uniformly along the isochrone
    (i.e., the initial mass function, $P(\gamma)$, is uniform) as well as with
    an intrinsic uniform 0.2~dex scatter orthogonal to the isochrone track. The
    isochrone is shifted by the stream's parallax modulus, which is derived from
    the astrometric distance track.  The background photometric coordinates are
    generated as a 2-dimensional Gaussian with non-zero covariance.  Thus, the
    synthetic data scenario described here generates a mock stream in
    astrometric coordinates and photometric coordinates, with density
    variations, a distance gradient which is consistent in parallax and distance
    modulus (i.e., magnitudes), all superimposed over a noisy field of
    background points in each astrometric and photometric dimension.

@@

    We pre-train a normalizing flow (see
    \autoref{sub:method:photometric_model:off_stream}) on an off-stream
    selection.  We train on the photometry.  You can see the selection in
    \autoref{fig:mock_data_photometric_background_selection}.  You can see both
    the on-stream and off-stream labels, blue and black respectively.  We select
    using an astrometric selection. In the photometry you can see that the
    on-stream selection has two populations, drawn from the isochrone and drawn
    from the background The off-stream selection, meanwhile, has only the
    background population. Next we train the full model.

>>>>

    We generate synthetic stream observations by sampling uniformly in $\phi_1$
    In $\phi_2$ we sample from a quadratic in $\phi_1$ with normally-distributed
    $0.15^{\circ}$ scatter.  The distances are sampled from a linear funciton in
    $\phi_1$ with normally-distributed $0.25 \ \unit{kpc}$ scatter, then
    transformed to the parallax $\parallax$.  Non-poissonian density variations
    are then simulated by evaluating the stream density with two Gaussian
    distributions centered at different values of $\phi_1$ with random
    amplitudes (which sum to $1$) and widths.  Stream stars are then removed
    using random sampling, weighted by their likelihood of belonging to the
    density. Post removal, the number of stream stars is
    $\variable{output/mock/nstream_variable.txt}$.
    $\variable{output/mock/nbackground_variable.txt}$ background astrometric points
    are generated from exponential distributions at each $\phi_1$.

    The synthetic stream's photometry is also simulated using a 12~Gyr, [Fe/H] =
    -1.35~dex NMIST isochrone \TODO{[cite brutus]}. The isochrone is truncated just
    short of the giant branch, and is sampled from uniformly along the isochrone
    (i.e., the initial mass function, $P(\gamma)$, is uniform) as well as with
    an intrinsic uniform 0.2~dex scatter orthogonal to the isochrone track. The
    isochrone is shifted by the stream's parallax modulus, which is derived from
    the astrometric distance track.  The background photometric coordinates are
    generated as a 2-dimensional Gaussian with non-zero covariance.  Thus, the
    synthetic data scenario described here generates a mock stream in
    astrometric coordinates and photometric coordinates, with density
    variations, a distance gradient which is consistent in parallax and distance
    modulus (i.e., magnitudes), all superimposed over a noisy field of
    background points in each astrometric and photometric dimension.

@@

    We select on-stream and off-stream regions using the known $\phi_1, \phi_2$
    range of the mock stream. The upper panel of
    \autoref{fig:mock_data_photometric_background_selection} shows these
    selections blue and black, respectively. The lower panel shows the on and
    off-stream selections in photometric coordinates. In this space it is
    apparent that the on-stream region has two populations, drawn from the
    isochrone and from background points included in the simple $\phi_1, \phi_2$
    box. In contrast the off-stream region, has only the the background population.
    We use the off-stream population to train a conditional normalizing flow
    (see \autoref{sub:method:photometric_model:off_stream}) on the magnitudes
    $g$ and $r$ and conditioned on $\phi_1$. The normalizing flow is now a
    non-parametric distribution characterizing the photometric background,
    including its variation over $\phi_1$. For this mock stream the flexibility
    of the normalizing flow greatly exceeds necessity: the background is
    drawn from a $\phi_1$-invariant multivariate normal, essentially the base
    distribution of the normalizing flow. It is therefore no surprise that
    the trained normalizing flow provides an excellent characterization of
    the background. We will later see that the flow performs similarly well
    on complex distributions. Having trained the background photometric model
    we proceed to train the full model.
